MIMO Channel Shaping and Rate Maximization\\Using Beyond Diagonal RIS

### 2024-12-29 manuscript/main.tex
:dizzy: Convergence proof

% \title{Channel Shaping Using Beyond Diagonal RIS:\\Analysis, Optimization, and Enhanced Flexibility}
% \title{MIMO Channel Shaping and Rate Maximization via Beyond Diagonal RIS}
% \title{MIMO Channel Shaping Using Beyond Diagonal RIS: Singular Value Redistribution and Rate Maximization}
\title{Beyond Diagonal RIS: MIMO Channel Shaping and Rate Maximization}
\author{
	\IEEEauthorblockN{
		Yang Zhao, \IEEEmembership{Member, IEEE,}
		Hongyu Li, \IEEEmembership{Graduate Student Member, IEEE,}\\
		Massimo Franceschetti, \IEEEmembership{Fellow, IEEE,}
		and Bruno Clerckx, \IEEEmembership{Fellow, IEEE}
	}
	\thanks{
		Yang Zhao, Hongyu Li, and Bruno Clerckx are with the Department of Electrical and Electronic Engineering, Imperial College London, London SW7 2AZ, U.K. (e-mail: \{\href{mailto:yang.zhao18@imperial.ac.uk}{yang.zhao18}, \href{mailto:c.li21@imperial.ac.uk}{c.li21}, \href{mailto:b.clerckx@imperial.ac.uk}{b.clerckx}\}@imperial.ac.uk).

@@

	% wider dynamic range (with better tradeoff) of singular values and significant power and rate gains, especially in large-scale \gls{mimo} systems.

@@

	\emph{Notation:}
	Italic, bold lower-case, and bold upper-case letters indicate scalars, vectors and matrices, respectively.
	$\jmath$ denotes the imaginary unit.
	% $\mathbb{C}$ represents the set of complex numbers.
	$\mathbb{H}^{n \times n}$, $\mathbb{H}_+^{n \times n}$, and $\mathbb{U}^{n \times n}$ denote the set of $n \times n$ Hermitian, positive semi-definite, and unitary matrices, respectively.
	% $\mathbf{0}$ and $\mathbf{I}$ are the all-zero and identity matrices with appropriate size, respectively.
	$\Re\{\cdot\}$ takes the real part of a complex number.
	% $\tr(\cdot)$ and $\det(\cdot)$ evaluate the trace and determinant of a square matrix, respectively.
	$\diag(\cdot)$ constructs a square matrix with arguments on the main (block) diagonal and zeros elsewhere.
	$\sv(\cdot)$, $\ran(\cdot)$, and $\ker(\cdot)$ evaluate the singular values (in descending order), column space (range), and kernel of a matrix, respectively.
	% $\sv(\cdot)$ extracts the singular values of a matrix as a non-ascending vector.
	% $\card(\cdot)$ returns the cardinality of a set.
	$\conv(\cdot)$ returns the convex hull of arguments.
	$\lvert \cdot \rvert$, $\lVert \cdot \rVert$, and $\lVert \cdot \rVert _\mathrm{F}$ denote the absolute value, Euclidean norm, and Frobenius norm, respectively.
	$\sigma_n(\cdot)$ and $\lambda_n(\cdot)$ are the $n$-th largest singular value and eigenvalue, respectively.
	% $\boldsymbol{\sigma}(\cdot)$ and $\boldsymbol{\lambda}(\cdot)$ are the corresponding vectors.
	$(\cdot)^*$, $(\cdot)^\mathsf{T}$, $(\cdot)^\mathsf{H}$, $(\cdot)^\dagger$ $(\cdot)^{(r)}$, $(\cdot)^{\star}$ denote the conjugate, transpose, conjugate transpose (Hermitian), Moore-Penrose inverse, $r$-th iterated point, and stationary point, respectively.
	$[N]$ is a shortcut for $\{1,2,\ldots,N\}$.
	$(\cdot)_{[x:y]}$ is a shortcut for $(\cdot)_x,(\cdot)_{x+1},\ldots,(\cdot)_y$.
	$\odot$ denotes the element-wise (Hadamard) product.
	$\mathcal{N}_{\mathbb{C}}(\mathbf{0}, \mathbf{\Sigma})$ is the multivariate \gls{cscg} distribution with mean $\mathbf{0}$ and covariance $\mathbf{\Sigma}$.
	$\sim$ means ``distributed as''.
\end{section}

@@

	\begin{subsection}{MIMO Interference Channel}
		We also consider a \gls{bd}-\gls{ris} aided \gls{mimo} \gls{ic} of $K$ transceiver pairs where each transmitter and receiver has $N_\mathrm{T}$ and $N_\mathrm{R}$ antennas, respectively, and the \gls{bd}-\gls{ris} has $N_\mathrm{S}$ scattering elements.
		This configuration is denoted as $(N_\mathrm{T} \times N_\mathrm{S} \times N_\mathrm{R})^K$.
		Let $\mathbf{H}_\mathrm{D}^{[kj]} \in \mathbb{C}^{N_\mathrm{R} \times N_\mathrm{T}}$, $\mathbf{H}_\mathrm{B}^{[k]} \in \mathbb{C}^{N_\mathrm{R} \times N_\mathrm{S}}$, $\mathbf{H}_\mathrm{F}^{[j]} \in \mathbb{C}^{N_\mathrm{S} \times N_\mathrm{T}}$ denote the direct channel from transmitter $j$ to receiver $k$, the backward channel of receiver $k$, and the forward channel of transmitter $j$, respectively, where $(j,k) \in [K]^2$.
		Assume all transmitter-\gls{ris}-receiver paths share the same \gls{bd}-\gls{ris} scattering matrix $\mathbf{\Theta}$.
		The equivalent channel from transmitter $j$ to receiver $k$ is
		\begin{equation}
			\mathbf{H}^{[kj]} = \mathbf{H}_\mathrm{D}^{[kj]} + \mathbf{H}_\mathrm{B}^{[k]} \mathbf{\Theta} \mathbf{H}_\mathrm{F}^{[j]} = \mathbf{H}_\mathrm{D}^{[kj]} + \sum_g \mathbf{H}_{\mathrm{B},g}^{[k]} \mathbf{\Theta}_g \mathbf{H}_{\mathrm{F},g}^{[j]},
			\label{eq:channel_equivalent_ic}
		\end{equation}
		where $\mathbf{H}_{\mathrm{B},g}^{[k]} \in \mathbb{C}^{N_\mathrm{R} \times L}$ and $\mathbf{H}_{\mathrm{F},g}^{[j]} \in \mathbb{C}^{L \times N_\mathrm{T}}$ are associated with \gls{ris} group $g$, corresponding to the $(g{-}1)L{+}1$ to $gL$ columns of $\mathbf{H}_{\mathrm{B},g}^{[k]}$ and rows of $\mathbf{H}_{\mathrm{F},g}^{[j]}$, respectively,
	\end{subsection}
\end{section}

@@

		Inequality \eqref{iq:sv_rd_max} states that
		if $\mathbf{H}_\mathrm{B}$ and $\mathbf{H}_\mathrm{F}$ are at least rank $k$, then
		with a \gls{d}-\gls{ris} or \gls{bd}-\gls{ris} of sufficiently large $N_\mathrm{S}$,
		the $n$-th singular value of $\mathbf{H}$ can be enlarged to the $(n-k)$-th singular value of $\mathbf{T}$, or suppressed to the $n$-th singular value of $\mathbf{T}$.
		Moreover, the first $k$ channel singular values are unbounded above\footnote{The energy conservation law $\sum_n \sigma_n^2(\mathbf{H}) \le 1$ still has to be respected in all cases. This constraint is omitted in context for brevity.} while the last $k$ channel singular values can be suppressed to zero.
		A special case is Corollary \ref{co:los} for \gls{los} channel\footnote{A similar eigenvalue result has been derived for \gls{d}-\gls{ris} only \cite{Semmler2023}.}.

@@

		Corollary \ref{co:nd_capacity_snr_general} also suggests that the power- and rate-optimal scattering matrices \eqref{eq:ris_nd_power_max} and \eqref{eq:ris_nd_rate_max} coincide with each other when the direct channel is negligible and the \gls{bd}-\gls{ris} is fully-connected.
		When either condition is not satisfied, active and passive beamforming are coupled and the rate-optimal solution involves alternating optimization.
		However, the power-optimal \gls{ris} still provides for a low-complexity decoupled solution.
		The details will be discussed in Section~\ref{sc:rate}.

@@

		The latter generalizes the conjugate gradient methods to Riemannian manifolds and updates the solution by addition and retraction, constituting a zigzag path departing from and returning to the manifold.

@@

		It is applicable to a wide range of \gls{bd}-\gls{ris} design problems where the objective $f$ is smooth or convex non-smooth (not necessarily the symmetric gauge function \eqref{ob:shaping}) and the only constraint is group-wise unitarity \eqref{cn:shaping_unitary}.

@@

			\item \emph{Translate to the Riemannian gradient at $\mathbf{\Theta}_g^{(r)}$:} At point $\mathbf{\Theta}_g^{(r)}$, the Riemannian gradient gives the steepest ascent direction on the manifold. It lies in the tangent space of the manifold $\mathcal{T}_{\mathbf{\Theta}_g^{(r)}}\mathbb{U}^{L \times L} \triangleq \{\mathbf{M} \in \mathbb{C}^{L \times L} \mid \mathbf{M}^\mathsf{H} \mathbf{\Theta}_g^{(r)} + {\mathbf{\Theta}_g^{(r)\mathsf{H}}} \mathbf{M} = \mathbf{0}\}$ and is obtainable by projection:

@@

			\item \emph{Translate to the Riemannian gradient at the identity:} The Riemannian gradient should be translated back to the identity for exploiting the Lie algebra\footnote{Lie algebra refers to the tangent space of the Lie group at the identity element. A Lie group is simultaneously a continuous group and a differentiable manifold. In this example, $\mathbb{U}^{L \times L}$ formulates a Lie group and the corresponding Lie algebra consists of skew-Hermitian matrices $\mathfrak{u}(L) \triangleq \mathcal{T}_{\mathbf{I}}\mathbb{U}^{L \times L} = \{\mathbf{M} \in \mathbb{C}^{L \times L} \mid \mathbf{M}^\mathsf{H} + \mathbf{M} = \mathbf{0}\}$.}:

@@

			\item \emph{Determine the conjugate direction:} The conjugate direction is obtained over the Riemannian gradient and previous direction as

@@

		\begin{algorithm}[!t]
			\caption{Group-wise geodesic \gls{rcg}}
			\label{ag:rcg}
			\begin{algorithmic}[1]
				\Require $f(\mathbf{\Theta})$, $G$
				\Ensure $\mathbf{\Theta}^\star$
				\Initialize {$r \gets 0$, $\mathbf{\Theta}^{(0)}$}
				\Repeat
				\For {$g \gets 1$ to $G$}
				\State $\nabla_{\mathrm{E},g}^{(r)} \gets$ \eqref{eq:gradient_eucl}, $\tilde{\nabla}_{\mathrm{R},g}^{(r)} \gets$ \eqref{eq:gradient_riem_tran}
				\State $\tilde{\gamma}_g^{(r)} \gets$ \eqref{eq:conjugate_parm_geod}, $\mathbf{D}_g^{(r)} \gets$ \eqref{eq:conjugate_dirn_geod}
				\If {$\Re\bigl\{\tr({\mathbf{D}_g^{(r)\mathsf{H}}} \tilde{\nabla}_{\mathrm{R},g}^{(r)})\bigr\} < 0$} \Comment{Non-ascent}
				\State $\mathbf{D}_g^{(r)} \gets \tilde{\nabla}_{\mathrm{R},g}^{(r)}$
				\EndIf
				\State $\mu \gets 0.1$ \Comment{Doubling cheaper than halving}
				\State $\mathbf{G}_{\mathbf{\Theta}_g^{(r)}}(\mu) \gets$ \eqref{eq:geodesic_tran}
				\While {$\scriptstyle f(\mathbf{G}_{\mathbf{\Theta}_g^{(r)}}(2\mu)) - f(\mathbf{\Theta}_g^{(r)}) \ge \mu \cdot \tr(\mathbf{D}_g^{(r)} {\mathbf{D}_g^{(r)\mathsf{H}}}) / 2$} \label{ln:armijo_start}
				\State $\mu \gets 2 \mu$
				\EndWhile
				\While {$\scriptstyle f(\mathbf{G}_{\mathbf{\Theta}_g^{(r)}}(\mu)) - f(\mathbf{\Theta}_g^{(r)}) < \mu / 2 \cdot \tr(\mathbf{D}_g^{(r)} {\mathbf{D}_g^{(r)\mathsf{H}}}) / 2$}
				\State $\mu \gets \mu / 2$
				\EndWhile \label{ln:armijo_end}
				\State $\mathbf{\Theta}_g^{(r+1)} \gets$ \eqref{eq:geodesic_tran}
				\EndFor
				\State $r \gets r+1$
				\Until $\lvert f(\mathbf{\Theta}^{(r)}) - f(\mathbf{\Theta}^{(r-1)}) \rvert / f(\mathbf{\Theta}^{(r-1)}) \le \epsilon$
			\end{algorithmic}
		\end{algorithm}

		Algorithm~\ref{ag:rcg} summarizes the proposed group-wise geodesic \gls{rcg} method for group-connected \gls{bd}-\gls{ris} design when the objective is smooth or convex non-smooth and the only constraint is group-wise unitarity.
		Compared to the non-geodesic approach, it leverages Lie algebra to replace the add-then-retract update with a multiplicative update \eqref{eq:geodesic_tran} along the geodesics of the Stiefel manifold.
		This appropriate parameter space leads to faster convergence and easier step size tuning.
		Convergence to a local optimum is still guaranteed if not initialized at a stationary point.
		The group-wise updates can be performed in parallel to facilitate large-scale \gls{bd}-\gls{ris} design problems.
		One can also directly operate on $\mathbf{\Theta}$ and pinching (i.e., keeping the main block diagonal and nulling others) relevant expressions to accelerate the algorithm for a large $G$.

		We now analyze the computational complexity of solving singular value shaping problem \eqref{op:shaping} by Algorithm~\ref{ag:rcg}.
		To update each \gls{bd}-\gls{ris} group, \gls{svd} of $\mathbf{H}$ requires $\mathcal{O}(N N_\mathrm{T} N_\mathrm{R})$, Euclidean gradient \eqref{eq:shaping_subdiff} requires $\mathcal{O}\bigl(L N (N_\mathrm{T}+N_\mathrm{R}+L+1) \bigr)$, Riemannian gradient translation \eqref{eq:gradient_riem_tran} requires $\mathcal{O}(L^3)$, deviation parameter \eqref{eq:conjugate_parm_geod} and conjugate direction \eqref{eq:conjugate_dirn_geod} together require $\mathcal{O}(L^2)$, and matrix exponential \eqref{eq:geodesic_tran} requires $\mathcal{O}(L^3)$ operations \cite{Moler2003}.
		The overall complexity is thus $\mathcal{O}\bigl(I_\text{RCG} G \bigl(N N_\mathrm{T} N_\mathrm{R} + L N (N_\mathrm{T}+N_\mathrm{R}+L+1) + I_\text{BLS} L^3\bigr)\bigr)$, where $I_\text{RCG}$ and $I_\text{BLS}$ are the number of iterations for geodesic \gls{rcg} and backtracking line search (i.e., line \ref{ln:armijo_start}--\ref{ln:armijo_end} of Algorithm \ref{ag:rcg}), respectively.

		% TODO: compare O(D) and O(BD)

		% \begin{subsubsection}{Pareto Frontier Characterization}
		% 	\label{sc:pareto}
		% 	We now characterize the Pareto frontier of singular values of a general $N_\mathrm{T} \times N_\mathrm{S} \times N_\mathrm{R}$ channel \eqref{eq:channel} by maximizing their weighted sum
		% 	\begin{maxi!}
		% 	{\scriptstyle{\mathbf{\Theta}}}{\sum_{n=1}^N \rho_n \sigma_n(\mathbf{H})}{\label{op:pareto}}{\label{ob:pareto}}
		% 	\addConstraint{\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g=\mathbf{I},}{\quad \forall g,}{\label{cn:pareto_unitary}}
		% 	\end{maxi!}
		% 	where $\rho_n \ge 0$ is the weight of the $n$-th singular value.
		% 	% that can be positive, zero, or negative.
		% 	Varying $\{\rho_n\}_{n \in [N]}$ characterizes the Pareto frontier enclosing the singular value region such that problem \eqref{op:pareto} generalizes most singular value shaping problems.
		% 	It can be solved optimally by Algorithm~\ref{ag:rcg} with the Euclidean gradient given in Lemma \ref{lm:pareto}.
		% 	\begin{lemma}
		% 		\label{lm:pareto}
		% 		One Euclidean sub-gradient of \eqref{ob:pareto} with respect to \gls{bd}-\gls{ris} group $g$ is
		% 		\begin{equation}
		% 			\label{eq:gradient_eucl_pareto}
		% 			% \frac{\partial \sum_n \rho_n \sigma_n(\mathbf{H})}{\partial \mathbf{\Theta}_g^*} =
		% 			\mathbf{H}_{\mathrm{B},g}^\mathsf{H} \mathbf{U} \diag(\rho_1,\ldots,\rho_N) \mathbf{V}^\mathsf{H} \mathbf{H}_{\mathrm{F},g}^\mathsf{H} \in \frac{\partial \sum_{n=1}^N \rho_n \sigma_n(\mathbf{H})}{\partial \mathbf{\Theta}_g^*}
		% 		\end{equation}
		% 		where $\mathbf{U}$ and $\mathbf{V}$ are any left and right singular matrices of $\mathbf{H}$, respectively.
		% 	\end{lemma}
		% 	\begin{proof}
		% 		Please refer to Appendix~\ref{ap:pareto}.
		% 	\end{proof}
		% \end{subsubsection}
	\end{subsection}
\end{section}

>>>>

% \title{Channel Shaping Using Beyond Diagonal RIS:\\Analysis, Optimization, and Enhanced Flexibility}
\title{MIMO Channel Shaping and Rate Maximization\\Using Beyond Diagonal RIS}
% \title{MIMO Channel Shaping Using Beyond Diagonal RIS: Singular Value Redistribution and Rate Maximization}
% \title{Beyond Diagonal RIS: MIMO Channel Shaping and Rate Maximization}
\author{
	\IEEEauthorblockN{
		Yang Zhao, \IEEEmembership{Member, IEEE,}
		Hongyu Li, \IEEEmembership{Graduate Student Member, IEEE,}\\
		Massimo Franceschetti, \IEEEmembership{Fellow, IEEE,}
		and Bruno Clerckx, \IEEEmembership{Fellow, IEEE}
	}
	\thanks{
		Yang Zhao, Hongyu Li, and Bruno Clerckx are with the Department of Electrical and Electronic Engineering, Imperial College London, London SW7 2AZ, U.K. (e-mail: \{\href{mailto:yang.zhao18@imperial.ac.uk}{yang.zhao18}, \href{mailto:c.li21@imperial.ac.uk}{c.li21}, \href{mailto:b.clerckx@imperial.ac.uk}{b.clerckx}\}@imperial.ac.uk).

@@

	% wider dynamic range (with better trade-off) of singular values and significant power and rate gains, especially in large-scale \gls{mimo} systems.

@@

	\emph{Notation:}
	Italic, bold lower-case, and bold upper-case letters indicate scalars, vectors and matrices, respectively.
	$\jmath$ denotes the imaginary unit.
	% $\mathbb{C}$ represents the set of complex numbers.
	$\mathbb{H}^{n \times n}$, $\mathbb{H}_+^{n \times n}$, and $\mathbb{U}^{n \times n}$ denote the set of $n \times n$ Hermitian, positive semi-definite, and unitary matrices, respectively.
	% $\mathbf{0}$ and $\mathbf{I}$ are the all-zero and identity matrices with appropriate size, respectively.
	$\Re\{\cdot\}$ takes the real part of a complex number.
	% $\tr(\cdot)$ and $\det(\cdot)$ evaluate the trace and determinant of a square matrix, respectively.
	$\diag(\cdot)$ constructs a square matrix with arguments on the main (block) diagonal and zeros elsewhere.
	$\sv(\cdot)$, $\ran(\cdot)$, and $\ker(\cdot)$ evaluate the singular values (in descending order), column space (range), and kernel of a matrix, respectively.
	% $\sv(\cdot)$ extracts the singular values of a matrix as a non-ascending vector.
	% $\card(\cdot)$ returns the cardinality of a set.
	$\conv(\cdot)$ returns the convex hull of arguments.
	$\lvert \cdot \rvert$, $\lVert \cdot \rVert$, and $\lVert \cdot \rVert _\mathrm{F}$ denote the absolute value, Euclidean norm, and Frobenius norm, respectively.
	$\sigma_n(\cdot)$ and $\lambda_n(\cdot)$ are the $n$-th largest singular value and eigenvalue, respectively.
	% $\boldsymbol{\sigma}(\cdot)$ and $\boldsymbol{\lambda}(\cdot)$ are the corresponding vectors.
	$(\cdot)^*$, $(\cdot)^\mathsf{T}$, $(\cdot)^\mathsf{H}$, $(\cdot)^\dagger$, $(\cdot)^{\star}$ denote the conjugate, transpose, conjugate transpose (Hermitian), Moore-Penrose inverse, and stationary point, respectively.
	$[N]$ is a shortcut for $\{1,2,\ldots,N\}$.
	$(\cdot)_{[x:y]}$ is a shortcut for $(\cdot)_x,(\cdot)_{x+1},\ldots,(\cdot)_y$.
	$\odot$ denotes the element-wise (Hadamard) product.
	$\mathcal{N}_{\mathbb{C}}(\mathbf{0}, \mathbf{\Sigma})$ is the multivariate \gls{cscg} distribution with mean $\mathbf{0}$ and covariance $\mathbf{\Sigma}$.
	$\sim$ means ``distributed as''.
\end{section}

@@

	\begin{subsection}{MIMO Interference Channel}
		We also consider a \gls{bd}-\gls{ris} aided \gls{mimo} \gls{ic} of $K$ transceiver pairs where each transmitter and receiver has $N_\mathrm{T}$ and $N_\mathrm{R}$ antennas, respectively, and the \gls{bd}-\gls{ris} has $N_\mathrm{S}$ scattering elements.
		This configuration is denoted as $(N_\mathrm{T} \times N_\mathrm{S} \times N_\mathrm{R})^K$.
		Let $\mathbf{H}_\mathrm{D}^{(kj)} \in \mathbb{C}^{N_\mathrm{R} \times N_\mathrm{T}}$, $\mathbf{H}_\mathrm{B}^{(k)} \in \mathbb{C}^{N_\mathrm{R} \times N_\mathrm{S}}$, $\mathbf{H}_\mathrm{F}^{(j)} \in \mathbb{C}^{N_\mathrm{S} \times N_\mathrm{T}}$ denote the direct channel from transmitter $j$ to receiver $k$, the backward channel of receiver $k$, and the forward channel of transmitter $j$, respectively, where $(j,k) \in [K]^2$.
		Assume all transmitter-\gls{ris}-receiver paths share the same \gls{bd}-\gls{ris} scattering matrix $\mathbf{\Theta}$.
		The equivalent channel from transmitter $j$ to receiver $k$ is
		\begin{equation}
			\label{eq:channel_interference}
			\mathbf{H}^{(kj)} = \mathbf{H}_\mathrm{D}^{(kj)} + \mathbf{H}_\mathrm{B}^{(k)} \mathbf{\Theta} \mathbf{H}_\mathrm{F}^{(j)} = \mathbf{H}_\mathrm{D}^{(kj)} + \sum_g \mathbf{H}_{\mathrm{B},g}^{(k)} \mathbf{\Theta}_g \mathbf{H}_{\mathrm{F},g}^{(j)},
		\end{equation}
		where $\mathbf{H}_{\mathrm{B},g}^{(k)} \in \mathbb{C}^{N_\mathrm{R} \times L}$ and $\mathbf{H}_{\mathrm{F},g}^{(j)} \in \mathbb{C}^{L \times N_\mathrm{T}}$ are associated with \gls{ris} group $g$, corresponding to the $(g{-}1)L{+}1$ to $gL$ columns of $\mathbf{H}_{\mathrm{B},g}^{(k)}$ and rows of $\mathbf{H}_{\mathrm{F},g}^{(j)}$, respectively,
	\end{subsection}
\end{section}

@@

		Inequality \eqref{iq:sv_rd_max} states that
		if $\mathbf{H}_\mathrm{B}$ and $\mathbf{H}_\mathrm{F}$ are at least rank $k$, then
		with a \gls{d}-\gls{ris} or \gls{bd}-\gls{ris} of sufficiently large $N_\mathrm{S}$,
		the $n$-th singular value of $\mathbf{H}$ can be enlarged to the $(n-k)$-th singular value of $\mathbf{T}$, or suppressed to the $n$-th singular value of $\mathbf{T}$.
		Moreover, the first $k$ channel singular values are unbounded above\footnote{The energy conservation law $\sum_{n=1}^N \sigma_n^2(\mathbf{H}) \le 1$ still has to be respected in all cases. This constraint is omitted in context for brevity.} while the last $k$ channel singular values can be suppressed to zero.
		A special case is Corollary \ref{co:los} for \gls{los} channel\footnote{A similar eigenvalue result has been derived for \gls{d}-\gls{ris} only \cite{Semmler2023}.}.

@@

		Corollary \ref{co:nd_capacity_snr_general} also suggests that the power- and rate-optimal scattering matrices \eqref{eq:ris_nd_power_max} and \eqref{eq:ris_nd_rate_max} coincide with each other when the direct channel is negligible and the \gls{bd}-\gls{ris} is fully-connected.
		When either condition is not satisfied, active and passive beamforming are coupled and the rate-optimal solution involves alternating optimization.
		However, the power-optimal \gls{ris} still provides for a low-complexity decoupled solution.
		The details will be discussed in Section~\ref{sc:rate_pc}.

@@

		The latter generalizes the conjugate gradient methods to Riemannian manifolds and updates the solution by addition and retraction, which constitutes a zigzag path departing from and returning to the manifold.

@@

		It is applicable to a wide range of \gls{bd}-\gls{ris} design problems where the objective $f$ is smooth or convex non-smooth\footnote{$f$ is not necessarily the symmetric gauge function \eqref{ob:shaping}.} and the only constraint is group-wise unitarity \eqref{cn:shaping_unitary}.

@@

			\item \emph{Translate to the Riemannian (sub-)gradient at $\mathbf{\Theta}_g^{(r)}$:} At point $\mathbf{\Theta}_g^{(r)}$, the Riemannian (sub-)gradient gives the steepest ascent direction on the manifold. It lies in the tangent space of the manifold $\mathcal{T}_{\mathbf{\Theta}_g^{(r)}}\mathbb{U}^{L \times L} \triangleq \{\mathbf{M} \in \mathbb{C}^{L \times L} \mid \mathbf{M}^\mathsf{H} \mathbf{\Theta}_g^{(r)} + {\mathbf{\Theta}_g^{(r)\mathsf{H}}} \mathbf{M} = \mathbf{0}\}$ and is obtainable by projection:

@@

			\item \emph{Translate to the Riemannian (sub-)gradient at the identity:} The Riemannian (sub-)gradient should be translated back to the identity for exploiting the Lie algebra\footnote{Lie algebra refers to the tangent space of the Lie group at the identity element. A Lie group is simultaneously a continuous group and a differentiable manifold. In this example, $\mathbb{U}^{L \times L}$ formulates a Lie group and the corresponding Lie algebra consists of skew-Hermitian matrices $\mathfrak{u}(L) \triangleq \mathcal{T}_{\mathbf{I}}\mathbb{U}^{L \times L} = \{\mathbf{M} \in \mathbb{C}^{L \times L} \mid \mathbf{M}^\mathsf{H} + \mathbf{M} = \mathbf{0}\}$.}:

@@

			\item \emph{Determine the conjugate direction:} The conjugate direction is obtained over the Riemannian (sub-)gradient and previous direction as

@@

		\begin{algorithm}[!t]
			\caption{Group-wise geodesic \gls{rcg}}
			\label{ag:rcg}
			\begin{algorithmic}[1]
				\Require $f(\mathbf{\Theta})$, $G$
				\Ensure $\mathbf{\Theta}^\star$
				\Initialize {$r \gets 0$, $\mathbf{\Theta}^{(0)}$}
				\Repeat
				\For {$g \gets 1$ to $G$}
				\State $\nabla_{\mathrm{E},g}^{(r)} \gets$ \eqref{eq:gradient_eucl}, $\tilde{\nabla}_{\mathrm{R},g}^{(r)} \gets$ \eqref{eq:gradient_riem_tran}
				\State $\tilde{\gamma}_g^{(r)} \gets$ \eqref{eq:conjugate_parm_geod}, $\mathbf{D}_g^{(r)} \gets$ \eqref{eq:conjugate_dirn_geod}
				\If {$\Re\bigl\{\tr({\mathbf{D}_g^{(r)\mathsf{H}}} \tilde{\nabla}_{\mathrm{R},g}^{(r)})\bigr\} < 0$} \Comment{Not ascent}
				\State $\mathbf{D}_g^{(r)} \gets \tilde{\nabla}_{\mathrm{R},g}^{(r)}$
				\EndIf
				\State $\mu \gets 0.1$
				\State $\mathbf{G}_{\mathbf{\Theta}_g^{(r)}}(\mu) \gets$ \eqref{eq:geodesic_tran}
				\While {$ f\bigl(\mathbf{G}_{\mathbf{\Theta}_g^{(r)}}(2\mu)\bigr) - f(\mathbf{\Theta}_g^{(r)}) \ge \mu \cdot \frac{\tr(\mathbf{D}_g^{(r)} {\mathbf{D}_g^{(r)\mathsf{H}}})}{2}$}\label{ln:armijo_start}
				\State $\mu \gets 2 \mu$
				\EndWhile
				\While {$f\bigl(\mathbf{G}_{\mathbf{\Theta}_g^{(r)}}(\mu)\bigr) - f(\mathbf{\Theta}_g^{(r)}) < \frac{\mu}{2} \cdot \frac{\tr(\mathbf{D}_g^{(r)} {\mathbf{D}_g^{(r)\mathsf{H}}})}{2}$}
				\State $\mu \gets \mu / 2$
				\EndWhile \label{ln:armijo_end}
				\State $\mathbf{\Theta}_g^{(r+1)} \gets$ \eqref{eq:geodesic_tran}
				\EndFor
				\State $r \gets r+1$
				\Until $\lvert f(\mathbf{\Theta}^{(r)}) - f(\mathbf{\Theta}^{(r-1)}) \rvert / f(\mathbf{\Theta}^{(r-1)}) \le \epsilon$
			\end{algorithmic}
		\end{algorithm}

		Algorithm~\ref{ag:rcg} summarizes the introduced group-wise geodesic \gls{rcg} method.
		Compared to the non-geodesic approach, it leverages Lie algebra to replace the add-then-retract update with a multiplicative update \eqref{eq:geodesic_tran} along the geodesics of the Stiefel manifold.
		This appropriate parameter space leads to faster convergence and easier step size tuning.
		Convergence to a local optimum is still guaranteed if not initialized at a stationary point.
		The group-wise updates can be performed in parallel to facilitate large-scale \gls{bd}-\gls{ris} design problems.
		One can also directly operate on $\mathbf{\Theta}$ and pinching (i.e., keeping the main block diagonal and nulling others) relevant expressions to accelerate the algorithm for a large $G$.

		We now analyze the computational complexity of solving singular value shaping problem \eqref{op:shaping} by Algorithm~\ref{ag:rcg}.
		To update each \gls{bd}-\gls{ris} group, \gls{svd} of $\mathbf{H}$ requires $\mathcal{O}(N N_\mathrm{T} N_\mathrm{R})$, Euclidean sub-gradient \eqref{eq:shaping_subdiff} requires $\mathcal{O}\bigl(L N (N_\mathrm{T}+N_\mathrm{R}+L) \bigr)$, Riemannian sub-gradient translation \eqref{eq:gradient_riem_tran} requires $\mathcal{O}(L^3)$, deviation parameter \eqref{eq:conjugate_parm_geod} and conjugate direction \eqref{eq:conjugate_dirn_geod} together require $\mathcal{O}(L^2)$, and matrix exponential \eqref{eq:geodesic_tran} requires $\mathcal{O}(L^3)$ operations \cite{Moler2003}.
		The overall complexity is thus $\mathcal{O}\bigl(I_\text{RCG} G \bigl(N N_\mathrm{T} N_\mathrm{R} + L N (N_\mathrm{T}+N_\mathrm{R}+L) + I_\text{BLS} L^3\bigr)\bigr)$, where $I_\text{RCG}$ and $I_\text{BLS}$ are the number of iterations for geodesic \gls{rcg} and backtracking line search (i.e., line \ref{ln:armijo_start}--\ref{ln:armijo_end} of Algorithm \ref{ag:rcg}), respectively.
		That is, $\mathcal{O}_\mathrm{D}\bigl(N_\mathrm{S}\bigr)$ for \gls{d}-\gls{ris} and $\mathcal{O}_\mathrm{BD}\bigl(N_\mathrm{S}^3\bigr)$ for fully-connected \gls{bd}-\gls{ris}.
		% That is, $\mathcal{O}_\mathrm{D}\bigl(I_\text{RCG} N_\mathrm{S} \bigl(N N_\mathrm{T} N_\mathrm{R} + N (N_\mathrm{T}+N_\mathrm{R}) + I_\text{BLS}\bigr)\bigr)$ for \gls{d}-\gls{ris} and $\mathcal{O}_\mathrm{BD}\bigl(I_\text{RCG} \bigl(N N_\mathrm{T} N_\mathrm{R} + N_\mathrm{S} N (N_\mathrm{T}+N_\mathrm{R}+N_\mathrm{S}) + I_\text{BLS} N_\mathrm{S}^3\bigr)\bigr)$ for fully-connected \gls{bd}-\gls{ris}.

		To validate Algorithm~\ref{ag:rcg} and quantify the shaping capability of \gls{bd}-\gls{ris}, we characterize the achievable singular value region of \gls{bd}-\gls{ris}-aided \gls{mimo} channel by considering the Pareto optimization problem
		\begin{maxi!}
			{\scriptstyle{\mathbf{\Theta}}}{\sum_{n=1}^N \rho_n \sigma_n(\mathbf{H})}{\label{op:pareto}}{\label{ob:pareto}}
			\addConstraint{\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g=\mathbf{I},}{\quad \forall g,}{\label{cn:pareto_unitary}}
		\end{maxi!}
		where $\rho_n \ge 0$ is the weight associated with the $n$-th channel singular value.
		Varying those weights help to characterize the Pareto frontier of singular values that encloses the achievable region.
		While the objective \eqref{ob:pareto} does not suggest any meaningful performance metric (e.g., capacity or power gain), there exists an implicit relation since stronger channel shaping capability translates to better wireless performance.
		Problem \eqref{op:pareto} also generalizes the \gls{dof} problem in Proposition \ref{pp:dof} and the individual singular value shaping problem in Corollary \ref{co:nd_sv_indl} and Proposition \ref{pp:rd}.
		It can be solved optimally by Algorithm~\ref{ag:rcg} with $\mathbf{D}_{[n,n]} = \rho_n$ in \eqref{eq:shaping_subdiff}.
	\end{subsection}
\end{section}

	\begin{section}{Rate Maximization}
		\label{sc:rate}
		In this section, we first solve rate maximization problem for \gls{bd}-\gls{ris}-aided \gls{mimo} point-to-point and interference channels optimally by joint beamforming design, and then exploit channel shaping for a low-complexity two-stage solution.
